---
title: "PMTCT variables and their associations in HIV study"
author: "Dolleen Osundwa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
The data collected is based on a research study that is aimed to follow up on HIV positive mothers for a duration of time (ante-natal and six months post-natal) monitoring their viral load, adherence to clinic visitations and their consistent uptake of ARV drugs. 
This analysis is aimed at having an overview of the relationship between some indicators that were focused on during the study.

```{r required_packages, echo=FALSE,comment=""}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(skimr))
suppressPackageStartupMessages(library(plotly))
library(plotly)
library(gganimate)
library(gifski)
library(readxl)
library(forcats)
library(ggplot2)
```

##Load Packages

The following R packages are useful for this analysis:

<ol>
<li> `readxl`
<li> `tidyverse`
<li> `skimr`
<li> `forcats`
<li> `ggplot2`
<li> `gganimate`
<li> `gifski`
<li> `plotly`
</ol>

## Load data files


```{r read_file,comment=""}
hiv <- read_xlsx("AmbiraFacilityData_HIV .xlsx",col_names = TRUE, na = "NA")

str(hiv)
names(hiv)
dim(hiv)
```

## Data Wrangling
The column names should be changed for easier analysis. Most of the columns' class are not in the proper form. The last column will also be removed since it has no use.

## Format the names of the columns

```{r mnames, comment = ""}

#Format the column names
new_names = c("consent", "pid", "entered_date", "age", "ccc_num", "art_regimen", "disclosure_status", "marital_status", "planned_pregnancy", "family_planning", "lmp", "edd", "delivery_date", "delivery_outcome", "baseline_vl", "postpartum_vl", "infant_num", "birth_weight", "infant_sex", "six_week_pcr", "six_month_pcr", "infant_art", "adverse_event", "study_status", "last")

hiv <- hiv %>% set_names(new_names)
```

## Drop the last column
```{r rm_col, comment=""}
## Remove the last column
hiv <- hiv %>% select(-last)

dim(hiv)

#View details
skimr::skim(hiv)
```
##Missing Values
Observe the level of missing values. Remove all missing values if the level is low.

```{r missingness, comment = ""}

#Compute missing values
sum(is.na(hiv))

#Available values
sum(!is.na(hiv))

#Total values
dim(hiv)
d1 <- dim(hiv)[1]
d2 <- dim(hiv)[2]
d1*d2

#Total values (another approach)
sum(sum(is.na(hiv)),sum(!is.na(hiv)))

#Proportion of missing values
mean(is.na(hiv))
```
##Change the class of columns

Some of the columns need type casting

```{r change_cols, comment = ""}

#Change the columns from character to factor
hiv$art_regimen <- factor(hiv$art_regimen)
hiv$disclosure_status <- factor(hiv$disclosure_status)
hiv$marital_status <- factor(hiv$marital_status)
hiv$planned_pregnancy <- factor(hiv$planned_pregnancy)
hiv$family_planning <- factor(hiv$family_planning)
hiv$delivery_outcome <- factor(hiv$delivery_outcome)
```

## Wrangling

```{r clean, comment = ""}

#Observe unique values for delivery outcome 
unique(hiv$delivery_outcome)

#Count values
hiv %>% select(delivery_outcome) %>% count(delivery_outcome)

#Change the necessary values
hiv <- hiv %>% mutate(deliveryOutcome2 = recode(hiv$delivery_outcome, "LIVE BIRTHS" = "LIVE BIRTH","TWINS" = "LIVE BIRTH"))

#Verify
hiv %>% select(deliveryOutcome2) %>% count(deliveryOutcome2)

#Unique values for infant sex
unique(hiv$infant_sex)

#Count
hiv %>% select(infant_sex) %>% count(infant_sex)

#Replace values
hiv <- hiv %>% mutate(infantSex2 = recode(hiv$infant_sex,"FEMALE        MALE" = "FEMALE","MALE             MALE" = "MALE"))

#Verify
hiv %>% select(infantSex2) %>% count(infantSex2)

#Pregnancy
unique(hiv$planned_pregnancy)

#Replace values
hiv <- hiv %>% mutate(planPreg = recode(hiv$planned_pregnancy, "YES                      YES" = "YES"))

```
##visualization

```{r viz, comment = ""}

#Check the range of the age to determine bins range 
(hiv$age);mean(hiv$age)

#Distribution of age with marital status
age_hist <- ggplot(hiv, aes(age, fill = marital_status)) + geom_histogram()

ageh <- ggplotly(age_hist)
ageh

#HIV status disclosure with marital status
bars <- ggplot(hiv, aes(x = disclosure_status, fill = marital_status)) + geom_bar() + transition_states(marital_status)

bars2 <-  bars + enter_fade() + exit_fade()
bars2

#Disclosure, age and marital status
bp <-  ggplot(hiv, aes(x = disclosure_status, y = age, fill = marital_status)) + geom_boxplot() + transition_states(marital_status)
bp

#Subset the column for conraception
fp2 <- hiv %>% select(family_planning) %>% count(family_planning)

#Pictorial count of family planning
contraceptives <- ggplot(fp2, aes(family_planning, n, fill = family_planning)) + geom_col() + geom_text(aes(label = n, vjust = -0.5)) + theme_minimal() + theme(legend.position = "none", axis.title.y = element_text(angle = 0)) + labs(title = "Histogram pf Types of Contraceptives")
contraceptives

#Family planning and pregnancy
pp <- ggplot(data = hiv, aes(x = planPreg, fill = family_planning)) + geom_bar(position = "dodge") +facet_wrap(~family_planning)
ggplotly(pp)

```

The average age is 30 years and the median is 32. From the data, at the average age of 30 years all participants were already married.
